---
title: "project"
author: "Aditya Thakur"
date: "11/04/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Loading packages:* <--- needs updating 

Apart from the tidyverse package that we have discussed in class, there was 
1 other package used for data wrangling: readxl. This package was important 
as our data was in a .xlsx file and so this package was important for the 
following data loading step. 

```{r}
library(tidyverse)
library(readxl)
library(lubridate)
library(GGally)
library(corrplot)
library(leaps)
```

*First Dataset:*

In this step, we loaded our primary dataset into R. This is an exhaustive 
dataset, consisting of 378,576 rows and 65 columns. First and foremost, it was
important to clean this data to select relevant information. 

```{r}
WDIEXCEL <- read_excel("data/WDIEXCEL 2.xlsx")

```

*Data Wrangling (first dataset):*

As a group, we picked a list of econometrics which we believed would allow us to
perform in-depth analysis. Our primary objective was to compare these metrics 
in developed and developing countries. Conveniently, this dataset contained 
2 values for the "Country Name" column which were: "Heavily indebted poor 
countries (HIPC)" and "OECD members". Instead of aggregating statistics from a
group of countries, we decided to use these previously created groups. 

```{r}
countries <-c("Heavily indebted poor countries (HIPC)","OECD members")
```

Next, we filtered the dataset to include only the metrics important to our 
study.

```{r}
econ_var<-c("GDP per capita (current US$)",
            "Gross domestic savings (current US$)",
            "Government expenditure on education, total (% of GDP)",
            "Profit tax (% of commercial profits)",
            "GDP growth (annual %)",
            "Foreign direct investment, net inflows (BoP, current US$)",
            "Foreign direct investment, net outflows (BoP, current US$)",
            "Unemployment, female (% of female labor force) 
            (modeled ILO estimate)",
            "Unemployment, male (% of male labor force) (modeled ILO estimate)")


WDIEXCEL<-WDIEXCEL %>% 
  filter(`Indicator Name` %in% econ_var,`Country Name` %in% countries)
```

*Missing Values:*

Dealing with missing values was a challenge with this dataset. Removing all of 
the missing values would compromise our analysis, since the missing values were 
scattered throughout the data set, sometime spanning across many different year 
and metrics. 

Upon further observation of the dataset, one can see that there 
are a few metrics where the majority of years contain empty measurements. For
example, a metric such as Profit Tax has just 1 positive measurement, in year 
2020. 

Similarly, Government Expenditure on Education too has very few positive 
values. Due to the large scale of missing values for these 2 groups, we decided
to exclude these 2 metrics from our analysis. Any attempt to extrapolate current
data to forecast for past years for these metrics would be highly unstable due
to the large scale of missing values. 

```{r}
WDIEXCEL <- WDIEXCEL[-c(5, 7,13, 15), ]
```

*Second Dataset:*

Our first dataset focused primarily on macroeconomic data pertaining to matters
of Gross Domestic Product (GDP), Foreign Direct Investment (FDI) and 
Unemployment among others. 

In order to gain a more holistic view of the different circumstances in the 
different sets fo countries, we decided to incorporate a dataset that focuses 
on individual health; matters such as: Mortality Rate, Education and others. 

We first imported the .xlsx file and similar to the first dataset, it was long 
and exhaustive. We once again chose a list of covariates on which we could 
filter the dataset.

```{r}
Health_data<-read_excel(
  "data/Data_Extract_From_Health_Nutrition_and_Population_Statistics.xlsx")

health_var<-c("Number of infant deaths",
              "People using at least basic drinking water services 
              (% of population)",
              "Mortality rate, under-5 (per 1,000)",
              "GNI per capita, Atlas method (current US$)",
              "Rural population (% of total population)",
              "School enrollment, primary (% gross)",
              "Unemployment, male (% of male labor force)")

Health_data<-Health_data %>% filter(`Series Name` %in% health_var)

```

Unlike the first dataset, the columns in this dataset were not easily 
accessible. They were in a format which made accessing yearly data difficult. 
In order to solve this, we used the column names from the first dataset (since
the timespan for both datasets was identical) and copied it.

```{r}
colnames(Health_data)[5:65]<-c(colnames(WDIEXCEL)[5:65])
```

To make the 1st and 2nd dataset more accessible, we decided to use pivot_longer 
and create a column for the years through which the dataset spans. We first 
gathered all column names corresponding to the years (columns 5 to 65) and 
then used the pivot_longer function:

```{r}
years<-c(colnames(WDIEXCEL)[5:65])

WDIEXCEL<- WDIEXCEL %>% 
  pivot_longer(years,names_to="year",values_to="count")
Health_data<-Health_data %>% 
  pivot_longer(years,names_to="year",values_to="count")
```

Furthermore, to make the process of combining both datasets simpler, we made
sure all identical columns had identical names.

```{r}
colnames(Health_data)[which(names(Health_data)=="Series Name")] <- 
  "Indicator Name"
colnames(Health_data)[which(names(Health_data)=="Series Code")] <- 
  "Indicator Code"
```

*Missing Values (timespan):*

For the WDI dataset, we removed particular metrics due to a lack of sufficient 
data. While there are some metrics that have comprehensive data, most do not 
have sufficient measurements for years prior to 1990. This is a trend that is 
noticable in both datasets. Therefore, we decided to limit our analysis to 
1991-2019. 2020 was excluded due to the lack of data. 

```{r}
WDIEXCEL<-WDIEXCEL %>% filter(year>=1991)
Health_data<-Health_data %>% filter(year>=1991)
```

*Combining datasets:*

Finally, we wanted to combine both datasets.Since we wanted to keep all metrics 
in both datasets, we used 'rbind'.

```{r}
econi<-rbind(WDIEXCEL,Health_data)
econi<-econi %>% 
  group_by(`year`) %>% 
  summarize(`Indicator Name`,`Country Name`,count) %>% 
  filter(!(year=="2020"))
```

*Cleaning and Transformation:*

Now that we have 1 datasets containing the relevant countries and econometrics 
spanning across years, we can now move on to the next step of transforming data 
and getting ready for analysis. 

```{r}
econi
#Modifying Columns
econi$year<-as.Date(econi$year,format="%Y")
econi$year<-year(ymd(econi$year))
econi$count<-as.numeric(econi$count)
econi

econi_wide<-econi %>% 
  group_by(`Indicator Name`) %>% 
  mutate(row=row_number()) %>% 
  pivot_wider(names_from = `Indicator Name`,values_from=count) %>% 
  select(-c(row))
```



